# Description

The dataset ${rwl}$ comprises measurements of $N=234$ tree ring sequences originating from the Bernese Oberland. The individual sequences are dated (the most recent ring has a year associated to it), which allows to give the timespan from oldest tree (first ring measured) to most recent tree as 1332 to 2017. The length of the series is therefore $T=686$ years.

Let $R_t = \{r_{ti}, i=1,...N\}$ the set of ring width measurements for fixed year $t, t \in [1332,2017]$.

Let $n_t = |R_t| = \sum^N_{i=1} \mathds{1}_{(y_{ti} \neq \text{NA})}$, the cardinality for a fixed year $t$, also denoted the *sample depth*.

Let $\bar{r_t} = \frac{1}{n_t}\sum^N_{i=1}r_{ti}*\mathds{1}_{(r_i \neq \text{NA})}$, the mean ring width for a fixed year $t$.

Let ${sd}_t = \sqrt{\frac{1}{n_t-1}\sum^N_{i=1}(y_i-x_t)^2*\mathds{1}_{(y_i \neq \text{NA})}}$, the sample standard deviation for year $t$.

Let $\bar{R}=\{\bar{r_t}, t=1,...,T\}$ the set of yearly mean tree ring widths of all the individual trees with data for a given year. The unit is `1/100 mm`.

```{r originalseries, fig.cap="The timeseries of yearly mean tree ring widths."}
seriesMeanPlot
```
The series of mean tree ring widths is shown in figure \ref{fig:originalseries}. It is obvious that there is no seasonal component in the data.
```{r datacharacteristics, fig.cap="Cardinality (left) and residuals after linear trend removal (right)",fig.width=7, out.width = "\\linewidth"}
grid.arrange(seriesDepthPlot,seriesHeteroskedasticityPlot,ncol=2)
```

The supplier of the data recommended to ignore data after 1800 due to insufficient quality of data for these years, with a prevalence of young trees distorting $\bar{r_t}$ in that time range (compare \ref{fig:originalseries}). Additionally, it was decided to discard data before 1400 because of a very small sample depth in years before 1400 (see figure \ref{fig:datacharacteristics}, left) . The analysis has therefore been conducted with the set $\bar{R^*}={r_t, t \in [1400;1800]}$.
