# Data {#data}
The dataset used for the analysis is supplied from the the archaeological service of the canton Berne through a personal contact at the dendrology laboratory in Sutz, BE, who also detailed the process of acquiring the samples (see also [@dendroBern]). The chronology comprises 234 tree ring sequences from different buildings in the Bernese Oberland that are all located above 900 meters. Typically, the wooden structure of old buildings is sampled by boring the beams. The drilling core is then analyzed by microscopy to record the sequence of tree ring widths. Spruce was a common building material and there is a wealth of houses supplying data. The data can be assumed to originate almost entirely from trees grown above 900 meters because trees were only very rarely transported uphill. 

We now give some definitions and show in a more rigorous manner how the mean and standard deviation are calculated from the raw dataset. Let be $R_t := \{r_{ti}, i=1,...N\}$ the set of ring width measurements $r$ for a fixed year $t$. Then, the cardinality for a fixed year $t$, also denoted the *sample depth*, is defined as $n_t := |R_t| = \sum^N_{i=1} \mathds{1}_{(y_{ti} \neq \text{NA})}$. The average ring width of a fixed year $t$ is then $\bar{r_t} := \frac{1}{n_t}\sum^N_{i=1}r_{ti} \mathds{1}_{(r_i \neq \text{NA})}$, and the sample standard deviation can be computed as ${sd}_t := \sqrt{\frac{1}{n_t-1}\sum^N_{i=1}(r_{it}-\bar{r_t})^2 \mathds{1}_{(y_i \neq \text{NA})}}$. Finally, the time series of yearly mean tree ring widths calculated from individual trees with data is defined by $\bar{R}:=\{\bar{r_t}, t=1,...,T\}$.

```{r originalseries, fig.cap="The timeseries of yearly mean tree ring widths. The grey overlay shows the sample depth for each year.", out.width="\\linewidth"}
plot(chron(rwl_df),ylab="Mean ring width [1/100 mm]")
```

The series of mean tree ring widths $\{\bar{r_t}\}$ is shown in Figure \ref{fig:originalseries}^[The plot is created with the `dplR` package.]. The gray overlay represents the sample depth $n_t$. From the plot, a correlation between sample depth $n_t$ and the average ring width $\bar{r_t}$ could be assumed. However, no significant relationship was found when we used the sample depth $n_t$ as an exogeneous variable in a linear model. In contrast to that, the Spearman's correlation test is significant at $\alpha=5\%$ with a p-value of $2.194e-06$ and estimated $\rho=`r cor.test(rwl_mean,rwl_depth,method="spearman",exact = F)$estimate`$. Transforming the dendrochronologies with respect to the sample depth $n_t$ could be an intresting approach for a future analysis, but is not part of this report. 

Furthermore, bigger climatic trends such as the small ice age, a period of low temperatures throughout Europe from the 15th to 18th century can be seen in the data. The overall decreasing trend could be explained by population growth and an intensifying forest cultivation practice. Cutting out large amounts of old trees from a forest leaves the young trees competing for sunlight. As vertical growth is more important during the first years of a tree's life, the year rings are generally smaller.

As recommended by the supplier of the data, we ignore data after 1800 due to the prevalence of young trees distorting average ring widths $\bar{r_t}$ in that time range. Additionally, we discard data before 1400 because of a very small sample depth in the years before. The unit of the untransformed ring width is 1/100 mm throughout the paper.

```{r fitsampledepth, results='markup', size="footnotesize", include=FALSE}
summary(lm(rwl_mean~year+rwl_depth))$coef
```
