---
title: "Dendrochronologies: Dating wooden artifacts"
author: "Time Series Analysis"
date: '`r Sys.Date()`'
bibliography: final_project.bib
output:
    pdf_document:
        toc: true
        toc_depth: 2
        number_sections: true
        includes:
            in_header: latex/mystyles.sty
            before_body: latex/before_body.tex
        fig_caption: yes
        fig_height: 6
        fig_width: 7
        keep_tex: yes
        latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error=TRUE,
                      warning=FALSE,
                      message=FALSE,
                      echo=FALSE,
                      cache=FALSE,
                      dpi=300,
                      fig.width=7, # Default figure widths
                      fig.asp=0.618,
                      fig.align="center",
                      fig.path = "./figures/",
                      out.width = "0.7\\linewidth")
knitr::opts_knit$set(eval.after='fig.cap')
```

```{r dataprep, results='hide',echo=FALSE}
library(itsmr)
library(dplyr)
op <- par(no.readonly = F)

rwl <- readRDS("data/rwl_900+.Rds")
rwl_mean <- apply(rwl, 1, mean, na.rm=T)
rwl_sd <- apply(rwl,1,sd,na.rm=T)
depthfun <- function(x) sum(!is.na(x))
rwl_depth <- apply(rwl,1,depthfun)
time <- seq(1332,2017,1)

rwl_mean_ts <- ts(rwl_mean,end=2017)
```

# Abstract
words words words

# Introduction

Dendrochronology is an archaeological method used for dating artifacts. Chronologies are constructed by measuring the patterns of tree ring widths from individual wooden artifacts, then overlaying the data to build up a continuous series of mean yearly ring widths spanning hundreds to thousands of years. Chronologies are generally constructed for defined geographical regions as the environment (i.e the local climate, soil properties and terrain) has substantial impact on the growth of trees.

Newly measured artifacts are matched to existing, dated chronologies serving the double purpose of strengthening the chronology and dating the artifact in question.

## Timeseries used

The chronology used for this project has been compiled from 234 individual measurements of buildings in the Bernese Oberland which are all located above 900 meters. Typically, the wooden structure of old buildings is sampled by boring the beams. The drilling core is then analyzed by microscoping, recording the sequence of tree ring widths. Spruce was a common building material and there is a wealth of houses supplying data. The data can be assumed to originate almost entirely from trees that grew above 900 meters since trees were rarely transported uphill. The data has been supplied by the archaeological service of the canton Berne [@dendroBern] through a personal contact at the dendrology laboratory in Sutz, BE.

## Timeseries analysis applied to dendrochronologies

- WOOLLONS: Usually ARMA1,1 / AR1 processes

- dplR: AR2

- We will look at AR1, ARMA1,1 more closely

# Methods

The dataset used as well as the methods employed are detailed.

## Data
The dataset ${rwl}$ comprises measurements of $N=234$ tree ring sequences originating from the Bernese Oberland. The individual sequences are dated (the most recent ring has a year associated to it), which allows to give the timespan from oldest tree (first ring measured )to most recent tree as 1332 to 2017. The length of the series is therefore $T=686$ years.

Let $R_t$ the set ${r_{ti}, i=1,...N}$ of ring width measurements for fixed year $t, t \in [1332,2017]$.

Let $n_t = |R_t| = \sum^N_{i=1} \mathds{1}{y_{ti} \neq NA}$, the cardinality for a fixed year $t$, also denoted the *sample depth*.

Let $\bar{r_t} = \frac{1}{n_t}\sum^N_{i=1}r_{ti}*\mathds{1}{r_i \neq NA}$, the mean ring width for a fixed year $t$.

Let ${sd}_t = \sqrt{\frac{1}{n_t-1}\sum^N_{i=1}(y_i-x_t)^2*\mathds{1}{y_i \neq NA}}$, the sample standard deviation for year $t$.

Let $\{\bar{r_t}, t=1,...,T\}$ the set of yearly mean tree ring widths of all the individual trees with data for a given year. The unit is `1/100 mm`.

```{r datacharacteristics, fig.cap="Cardinality (left), mean ring width (center) and SD (right) per year", out.width = "\\linewidth"}
par(mfrow=c(1,3))
plot(rwl_depth~time,type="l",xlab="Year",ylab="Sample depth [#]")
plot(rwl_mean~time,type="l" ,xlab="Year",ylab="Mean ringwidth [1/100 mm]")
plot(rwl_sd~time,type="l",xlab="Year",ylab="SD [1/100 mm]")
par(op)
```

The supplier of the data recommended to ignore data after 1800 due to insufficient quality of data for these years, with a prevalence of young trees distorting $\bar{r_t}$ in that time range. Additionally, it was decided to discard data before 1400 because of a very small sample depth in years before 1400. The analysis has therefore been conducted with the set ${r_t, t \in [1400;1800]}$.

## Stationarity of the time series

In the following section, removal of trend and model fitting will be discussed. As seen in [fig:datacharacteristics], there is no apparent seasonal component present in the data.  

### Removing trend

Several approaches were studied to deal with heteroskedasticity and trend removal. These approaches wil be presented and discussed here. 

Overview of the methods considered:

1. **Woollons** Described in [@woollons1990time]. The proposed approach is an estimate of the form $Y = alpha*t^(beta)*exp(delta*t)$. Taking the log on both sides yields something linear in $ln(t)$
2. **Power tranformation of residuals**
  2.1 
  2.2
  2.3

3. **Box-Cox transform** Tranformation that yields a normally distributed timeseries
4. **Log transformation, polynomials** 

## imputing strategies -> Florian

    - imputeTS: na.kalman -> emphasize this
    
    - forecast: na.interp
    
    - zoo: na.approx
    
## Use of exogeneous data


# Results & Discussion

## Data -> David

- Transformation, model fitting 



## Imputation -> Florian

- Introduce missing values

- imputation using different strategies

- discussion of imputation results

# Conclusion

To be seen

# References

<div id="refs"></div>

\\clearpage

```{r child="appendix.Rmd" }
```
