# Modelling / Analysis

## Transformation


## Acieving stationarity

The approaches yielded very different results, as can be observed in figure \ref{fig:stationarity_alternatives}. The top plot shows the series that was used for further analyis. It is not immediately apparent from the plot, but the transformed series passed the stationarity tests(see \ref{fig:}). The method proposed by @woollons1990time had much larger variance and is therefore shown separately in figure \ref{fig:stationaritywoollons}. Power transform and Box-Cox transform shown at the bottom are effectively the inverse of each other.

```{r stationarity_alternatives, fig.cap="Comparison of different methods to achieve a stationary time series. Top: log-transformed with order 1 trend removal, bottom left: power transform and box-cox transform, bottom right: Woollons method.",out.width="\textwidth"}
layout <- rbind(c(1,1),c(2,3))
grid.arrange(stationarity_data_log_order1_varstab,
             stationarity_woollons_plot,
             stationarity_powerbox_plot
             , ncol=1)
```

## Model selection

The stationary time series labelled "Log trans., order1" in figure \ref{fig:stationaritycomparison} was chosen for further analysis as it appeared the most stationary (mean: `r Xt.bar`.  It's ACF / PACF plots are shown in figure \ref{fig:acfpacf_chosen}. The shape of the ACF and PACF give some basis to argue for either an AR(1) or ARMA(1,1) process.

```{r acfpacf_chosen, fig.cap="ACF (left) and PACF (right) of the chosen stationary time series."}
acf <- ggAcf(data_log_order1_varstab$y, main="") +
    theme(aspect.ratio = 0.618) # looks exponential
pacf <- ggPacf(data_log_order1_varstab$y,main="") +
    theme(aspect.ratio = 0.618) 
grid.arrange(acf,pacf,ncol=2)

```

Confirming the graphical analyis, an $ARMA(1,1)$ model was found through `autofit` from package `itsmr` with parameters `autofit(x=Xt,p=0:3, q=0:3,mean.include=F)`. We then cross-checked with the function `arima` and found similar parameters. Additionally, it was decided to also include an $AR(1)$ process. An overview of the parameters for each model and their standard deviation is given in table \ref{tab:modelcomp}.

```{r modelcomp}
kable(modelComparisonTable, booktabs=T,
      caption="Model parameters for ARMA(1,1) and AR(1).",
      escape=FALSE) %>%
kable_styling(latex_options = c("hold_position"), position = "center") %>%
group_rows("ARMA(1,1)",1,3) %>%
group_rows("AR(1)",4,5)
```
The validity of both models was assessed graphically through `tsidag`. For the $ARMA(1,1)$ model, the residuals are confirmed to be white noise. The Ljung-box statistic for various lags showed that for the $AR(1)$ model, the residuals are not white noise. The plots are shown in figure \ref{fig:tsdiag}.

```{r tsdiag, fig.cap="Diagnostic plots for the fitted models. Left: ARMA(1,1), right: AR(1).", out.width = "\\linewidth"}
grid.arrange(tsdiag(Xt.arma11),tsdiag(Xt.ar1),ncol=2)
```

## Imputation

- Introduce missing values

- imputation using different strategies

- discussion of imputation results
