# Results & Discussion

## Stationarity

The approaches yielded very different results, as can be observed in figure  \ref{fig:stationaritycomparison}. The method proposed by [@woollons1990time] had much larger variance and is therefore shown separately in figure \ref{fig:stationaritywoollons}.

```{r stationaritycomparison, fig.cap="Comparison of different methods to achieve a stationary time series."}
stationarity_comparison_plot
```

```{r stationaritywoollons, fig.cap="Timeseries tranformed according to [@woollons1990time]"}
stationarity_woollons_plot
```

The stationary time series labelled "Log trans., order1" in figure \ref{fig:stationaritycomparison} was chosen for further analysis. It's ACF / PACF plots are shown in figure \ref{fig:acfpacf_chosen}. The shape of both ACF and PACF give some basis to argue for either an AR(1) or ARMA(1,1) process.

The ACF / PACF plots for all approaches except for the variance-stabilized log transform with an order-1 trend indicated either non-stationarity or complex processes. These plots are shown in the appendix for reference.

```{r acfpacf_chosen, fig.cap="ACF (left) and PACF (right) of the chosen stationary time series."}

acf <- ggAcf(data_log_order1_varstab$y, main="") +
    theme(aspect.ratio = 0.618) # looks exponential
pacf <- ggPacf(data_log_order1_varstab$y,main="") +
    theme(aspect.ratio = 0.618) 
grid.arrange(acf,pacf,ncol=2)

```
## Considered processes: ARMA(1,1), AR(1), AR(2)

## Imputation

- Introduce missing values

- imputation using different strategies

- discussion of imputation results

## Exogeneous data