---
title: "Proposal"
author: "Time series_spruce_sub900 Analysis"
date: "8 März 2018"
bibliography: ../final_project.bib
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

Dendrochronology data has been supplied by the archaeological service of the canton Berne [@dendroBern] through a personal contact at the dendrology laboratory in Sutz, BE.

The chronologies have been compiled from individual measurements in the canton of Berne. Typically, the wooden structure of old buildings is sampled by boring the beams. The drilling core is then analyzed by microscoping, recording the sequence of tree ring widht. Subsequently, the sample is matched with existing chronological data, serving the double purpose of strengthening the chronology by overlaying a certain period with more evidence as well as dating the sample in an effort to date the construction year of the sampled building.

All chronologies end in 2017 and have varying lengths of at least 650 years.
The supplier of the data recommended ignoring data after 1800 due to an insufficient amount of data (individual treering measurements to compute a robust annual mean) for the period after 1800.

## Spruce data above 900 m
This chronology was compiled from buildings in the Bernese Oberland which are located above 900 m. Spruce was a common building material and there is a wealth of houses supplying data. It is almost sure that trees have not been transported uphill when building houses, so this data can be assumed to originate almost entirely from trees that grew above 900 m. It is known that treering width at high altitudes is mostly dependent on the average temperature during spring.

## Preliminary Analysis
```{r preparedata, echo=FALSE}
# Reading in the spruce data for altitudes below 900 m
spruce_sub900 <- read.table("../data/62163_picea_sub900.txt",
                                      skip=4,nrows=89)
spruce_sub900_lastrow <- read.table("../data/62163_picea_sub900.txt",
                                              skip=94,nrows=1)

# Spruce data above 900 m
spruce_sup900 <- read.table("../data/62167_picea_abv900.txt",
                            skip=4,nrows=68)
spruce_sup900_lastrow <- read.table("../data/62167_picea_abv900.txt",
                            skip=72,nrows=1)

ssub9 <- spruce_sub900[,-c(1,2,3)]
ssub9l <- spruce_sub900_lastrow[,-c(1,2,3)]

ssup9 <- spruce_sup900[,-c(1,2,3)]
ssup9l <- spruce_sup900_lastrow[,-c(1,2,3)]

series_spruce_sub900 <- c()
for (i in 1:nrow(ssub9)) {
    for(j in 1:ncol(ssub9)){
        series_spruce_sub900 <- c(series_spruce_sub900,as.integer(ssub9[i,j]))
    }
}
series_spruce_sub900 <- c(series_spruce_sub900,as.integer(ssub9l))

series_spruce_sup900 <- c()
for (i in 1:nrow(ssup9)) {
    for (j in 1:ncol(ssup9)) {
        series_spruce_sup900 <- c(series_spruce_sup900,as.integer(ssup9[i,j]))
    }    
}
series_spruce_sup900 <- c(series_spruce_sup900,as.integer(ssup9l))
# Some preliminary analyses
library(itsmr)

spruce_sub900_ts <- ts(series_spruce_sub900,start=1121,end=1800)
spruce_sup_900_ts <- ts(series_spruce_sup900, start=1332,end=1800)

```



```{r plotdata, fig.cap="Timeseries for spruce below / above 900 m"}
par(mfrow=c(1,2))
plot(spruce_sub900_ts,
     main="Spruce, < 900 m",
     ylab="tree ring width [1/100 mm]",
     xlab="year")

plot(spruce_sup_900_ts,
     main="Spruce, > 900 m",
     ylab="tree ring width [1/100 mm]",
     xlab="year")

```


### Spruce below 900 m
A first analysis of the time series for spruce below 900 m
```{r modelfitting}
t <- seq(from=1,length=length(spruce_sub900_ts))
model <- lm(spruce_sub900_ts~t)
summary(model)

resid_spruce_sub900 <- residuals(model)
resid_spruce_sub900_ts <- ts(resid_spruce_sub900,end=1800)
mean(resid_spruce_sub900_ts) # ~ zero mean
qqnorm(resid_spruce_sub900_ts) # close to a normal distribution
hist(resid_spruce_sub900_ts) # slightly right-skewed

plotc(resid_spruce_sub900_ts)

Box.test(resid_spruce_sub900_ts,lag=200,type="Ljung-Box")
# Reject H0: The residuals are not iid
# Which was to be expected, given the nature of the data

# Sample ACF shows non-stationarity.
acf(resid_spruce_sub900_ts,lag.max = 250)

# try to paramtrize an AR(1) model
ar1d1 <- resid_spruce_sub900[1:(length(resid_spruce_sub900)-1)]
ar1d2 <- resid_spruce_sub900[-1]
plot(ar1d2~ar1d1)
ar1_model <- lm(ar1d2~ar1d1)
summary(ar1_model) # 
acf(ar1_model$residuals,lag.max = 200) # already looks a lot better, but to many gamma(h) outside of range

# AR(2)?
ar2d1 <- resid_spruce_sub900[1:(length(resid_spruce_sub900)-2)]
ar2d2 <- resid_spruce_sub900[2:(length(resid_spruce_sub900)-1)]
ar2d3 <- resid_spruce_sub900[3:length(resid_spruce_sub900)]
ar2_model <- lm(ar2d3~ar2d1+ar2d2)
summary(ar2_model)
acf(ar2_model$residuals) # still not...

pacf(resid_spruce_sub900_ts,100)
pacf(ar2_model$residuals,100) #

# Include a longterm temperature trend as another explanatory variable

climate <- read.csv("../data/loehle_global_temp_reconstruction.csv",header=T)
climate_ts <- ts(climate$Temp..Anom.,start=16,end=1935)

climate_sub900 <- climate[climate$Year..AD.>=1121 & climate$Year..AD. <= 1800,]
climate_sup900 <- climate[climate$Year..AD.>=1332 & climate$Year..AD. <= 1800,]

climate_sub900_ts <- ts(climate_sub900$Temp..Anom.,start=1121,end=1800)
plot(spruce_sub900_ts,type="l")
lines(climate_sub900_ts*mean(spruce_sub900_ts),col="red",type="l")
#plotc(climate_sub900_ts)

model2 <- lm(spruce_sub900_ts~climate_sub900_ts+t)
summary(model2)
plotc(model2$residuals)

```

### Spruce above 900 m
A first analysis of the time series for spruce above 900 m
```{r modelfitting_sup900}
t <- seq(from=1,length=length(spruce_sup_900_ts))
model <- lm(spruce_sup_900_ts~t)
summary(model) # cubic trend is still significant

resid_spruce_sup900 <- residuals(model)
resid_spruce_sup900_ts <- ts(resid_spruce_sup900,end=1800)
mean(resid_spruce_sub900_ts) # ~ zero mean
qqnorm(resid_spruce_sup900_ts) # close to a normal distribution
hist(resid_spruce_sup900_ts) # slightly right-skewed

plotc(resid_spruce_sup900_ts)

Box.test(resid_spruce_sup900_ts,lag=200,type="Ljung-Box")
# Reject H0: The residuals are not iid
# Which was to be expected, given the nature of the data

# Sample ACF shows non-stationarity.
acf(resid_spruce_sup900_ts,lag.max = 250)

# try to paramtrize an AR(1) model
ar1d1 <- resid_spruce_sup900[1:(length(resid_spruce_sup900)-1)]
ar1d2 <- resid_spruce_sup900[-1]
plot(ar1d2~ar1d1)
ar1_model <- lm(ar1d2~ar1d1)
summary(ar1_model) # 
acf(ar1_model$residuals,lag.max = 200) # already looks a lot better, but to many gamma(h) outside of range

# AR(2)?
ar2d1 <- resid_spruce_sup900[1:(length(resid_spruce_sup900)-2)]
ar2d2 <- resid_spruce_sup900[2:(length(resid_spruce_sup900)-1)]
ar2d3 <- resid_spruce_sup900[3:length(resid_spruce_sup900)]
ar2_model <- lm(ar2d3~ar2d1+ar2d2)
summary(ar2_model)
acf(ar2_model$residuals) # still not...

pacf(resid_spruce_sup900_ts,100)
pacf(ar2_model$residuals,100) #


```

# Questions to be answered

Climatic trends are to be identified by analyzing the correlation between 3 different chronologies.

# Methods

The project team will answer the questions using the following methods:

* Method 1
* Method 2


# References
